---
title: "Explore the Fire Ignition Points Attribute Table"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

In this Notebook, we'll do compute some summary stats for the attribute table of the Yosemite fire ignition points, then use attribute queries to filter them.

# SETUP

Load the packages we'll need for this notebook (sf, dplyr, and tmap):

```{r chunk01}
library(sf)
library(dplyr)
library(tmap)
```

\

Pro-actively resolve any name clashes:

```{r chunk02}
library(conflicted)

# Set conflict preferences
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("arrange", "dplyr", quiet = TRUE)
```

\

Next import fire ignition points layer:

```{r chunk03}
## Define the location of the geodatabase
gdb_fires_fn <- "./data/yose_firehistory.gdb"
file.exists(gdb_fires_fn)

## View the layers in this source
st_layers(gdb_fires_fn)

## Import the historic fires ignition points
yose_fires_pt <- st_read(gdb_fires_fn, layer="YNP_FireHistoryPoints")

## Preview
glimpse(yose_fires_pt)
```

\

Import the park boundary and project it to UTM Zone 11N NAD83 (for plotting):

```{r chunk04}
epsg_utm11n_nad83 <- 26911

yose_bnd_sf <- st_read(dsn="./data", layer="yose_boundary") |> 
  st_transform(epsg_utm11n_nad83)
```

\

Plot them:

```{r chunk05}
tm_shape(yose_bnd_sf) +
  tm_borders() +
tm_shape(yose_fires_pt) + 
  tm_dots(col = "red", alpha = 0.5, size = 0.05) 
```

\

# DESCRIPTIVE STATS OF THE ATTRIBUTE TABLE

If you want to compute descriptive stats on the attribute table, without regard to the geometry, you can extract the attribute table as a data frame with `st_drop_geometry()`:

```{r chunk06}
yose_fires_pt |> st_drop_geometry() |> head()
```

\

## CHALLENGE QUESTIONS

Q01. What is the distribution of acres burned?

TIP: To view the distribution, compute the quantiles or make a histogram plot.

[Answer](http://bit.ly/3kXVHZA)

```{r chunk07}
## Because the 'ACRES' column is numeric, you can compute the quantitles with summary()
summary(yose_fires_pt$ACRES)

## Or:
quantile(yose_fires_pt$ACRES)

## Histogram
hist(yose_fires_pt$ACRES, breaks = 20)
```

\

Q02. Compute a frequency table of fire causes.

[Answer](http://bit.ly/3ZzvT5a)

```{r chunk08}
table(yose_fires_pt$CAUSE)

#Or 
yose_fires_pt |> st_drop_geometry() |> group_by(CAUSE) |> tally()
```

\

Q03. What was the largest fire?

[Answer](http://bit.ly/3Zx0K2k)

```{r chunk09}
yose_fires_pt |> st_drop_geometry() |> slice_max(ACRES, n=1)
```

\

Or:

```{r chunk10}
yose_fires_pt |> 
  st_drop_geometry() |> 
  arrange(desc(ACRES)) |> 
  slice(1)
```

\

Q04. For each decade, compute the number of fires, and the mean fire size.

[Answer](http://bit.ly/3F5Zb3e)

```{r chunk11}
yose_fires_pt |> 
  st_drop_geometry() |> 
  group_by(DECADE) |> 
  summarise(count = n(), avg_size_acres = mean(ACRES))
```

\

# ATTRIBUTE QUERIES

Attribute queries can be performed using dplyr `filter()`.

Example: make a plot of the fire ignition points from the 1930s:

```{r chunk12}
fires30s_sf <- yose_fires_pt |> 
  filter(DECADE == 1930)

head(fires30s_sf)

tm_shape(yose_bnd_sf) +
  tm_borders() +
tm_shape(fires30s_sf) + 
  tm_dots(col = "red", size = 0.2) 
```

\

## CHALLENGE QUESTIONS

Q05. Make a plot of all the human caused fire ignition points.

[Answer](http://bit.ly/3l1uLZ0)

```{r chunk13}
fires_hc_sf <- yose_fires_pt |>  
  filter(CAUSE == 'HC')

tm_shape(yose_bnd_sf) +
  tm_borders() +
tm_shape(fires_hc_sf) + 
  tm_dots(col = "red", size = 0.2)
```

\

Q06. Make a plot of all the human caused fire ignition points in the 1980s (only).

[Answer](http://bit.ly/3l05iiB)

```{r chunk14}
fires_hc80s_sf <- yose_fires_pt |>  
  filter(CAUSE == 'HC', DECADE == 1980)

tm_shape(yose_bnd_sf) +
  tm_borders() +
tm_shape(fires_hc80s_sf) + 
  tm_dots(col = "red", size = 0.2) +
tm_layout(main.title = "Human Caused Fires During the 1980s",
          main.title.size = 1)
```

\

Q07. Plot the *human caused fires each decade* (one plot per decade), for each decade from the 1980s onward:

Tip: See `tm_facets()`

[Answer](http://bit.ly/3mGs7YY)

```{r chunk15}
fires_hc50s20_sf <- yose_fires_pt |>  
  filter(CAUSE == 'HC', DECADE >= 1980)

tm_shape(yose_bnd_sf) +
  tm_borders() +
tm_shape(fires_hc50s20_sf) + 
  tm_dots(col = "red", size = 0.2) +
tm_facets(by = "DECADE") +
  tm_layout(main.title = "Human Caused Fires Per Decade",
            main.title.size = 1)
```

