---
title: "Spatial Queries: Find Yosemite POIs in the Upeer Merced Subbasin"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

```{css echo = FALSE}
h1,h2 {font-weight:bold;}
```


In this Notebook we'll use geoprocessing functions from `sf` to identify Yosemite Points-of-Interest that fall within the **Upper Merced Subbasin**. 

## Setup

Load the packages we'll need and set tmap mode to 'plot':

```{r chunk01, message = FALSE}
library(sf)
library(tmap)
tmap_mode("plot")
```

Load `dplyr` and set name conflict preferences:

```{r chunk02, message = FALSE}
library(dplyr)

## Load the conflicted package
library(conflicted)

# Set conflict preferences
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("arrange", "dplyr", quiet = TRUE)
```

\

# Practice Querying with Sample Data

## Import Practice Data

First we import some practice data:

```{r chunk03}
circles_sf <- st_read("./data/test_circles.geojson")
circles_sf

pts_sf <- st_read("./data/test_pts.geojson")
pts_sf
```

\

Plot the points on top of the circles:

```{r chunk04, warning = FALSE}
circles_sf |> 
  mutate(circle_color = c("red", "blue", "green")) |> 
  st_cast("MULTILINESTRING") |> 
  tm_shape() +
  tm_lines("circle_color") +
  tm_shape(circles_sf) +
  tm_text("circle_id") +
  tm_shape(pts_sf) +
  tm_dots(col = "dimgray") +
  tm_grid(labels.show = TRUE, lines = FALSE)
```

## Identify the points in Circle A

Next we identify the points in circle A using a spatial predicate function (st_intersects). We could also copy the points in Circle A with st_intersection(), but there are times when you don't need or want to make copies of the data.

```{r chunk05}
circle_a_sf <- circles_sf |> filter(circle_id == "A")

pt_in_circle_a_yn_mat <- pts_sf |> 
  st_intersects(circle_a_sf,
                sparse = FALSE)
  
head(pt_in_circle_a_yn_mat)
```

Since we have a column of TRUE/FALSE values, we can subset those features using the dplyr `filter` function:

```{r chunk06, warning = FALSE}
## Copy the points in circle A to a new object. Note in the filter expression
## we use square bracket notation to pull out the first column of the matrix 

a_pts_sf <- pts_sf |> 
  filter(pt_in_circle_a_yn_mat[,1])

## Plot to verify
tm_shape(circles_sf) +
  tm_borders() +
  tm_text("circle_id") +
tm_shape(a_pts_sf) +
  tm_dots(col = "red", size = 0.1) +
tm_grid(labels.show = TRUE, lines = FALSE)
```

\

## CHALLENGE: How many points in circle B?

[Answer](https://bit.ly/3xwk0xV)

```{r chunk07}
circle_b_sf <- circles_sf |>
  filter(circle_id == "B")

pts_sf |> 
  st_intersects(circle_b_sf, sparse = FALSE) |> 
  sum()
```

\

## CHALLENGE: Plot the points that fall within Circle B *and* Circle C

[Answer](https://bit.ly/3yvJk8H)

```{r chunk08}
pts_circle_bc_sf <- pts_sf |> 
  st_intersection(circles_sf |> filter(circle_id == "B")) |> 
  st_intersection(circles_sf |> filter(circle_id == "C"))

tm_shape(circles_sf) +
  tm_borders() +
  tm_text("circle_id") +
tm_shape(pts_circle_bc_sf) +
  tm_dots(col = "red", size = 0.1) +
tm_grid(labels.show = TRUE, lines = FALSE)
```

\

## CHALLENGE: How many points don't fall in any circle?

This is a little more challenging. One approach we could take is to test for intersection for all points and all circles. This will return a matrix of TRUE / FALSE values, with one row for each point and one column for each circle. If a row has three FALSE values, it means that point doesn't intersect any circle. 

[Answer](https://bit.ly/2VCUfz3)

```{r chunk09}
## Approach one: Test intersection for all circles
intersect_yn_mat <- pts_sf |> 
  st_intersects(circles_sf, sparse = FALSE)
head(intersect_yn_mat)

## Sum up the rows. If a the point doesn't intersect any circle, the sum of the row will be 0
outside_all_circles <- rowSums(intersect_yn_mat)

## Count the number of rows where the total is 0. These are the points that don't intersect any circle.
sum(outside_all_circles == 0)

## Plot to make sure
pts_outside_circles_sf <- pts_sf |> 
  filter(outside_all_circles == 0)

tm_shape(circles_sf) +
  tm_borders() +
  tm_text("circle_id") +
tm_shape(pts_outside_circles_sf) +
  tm_dots(col = "dimgray", size = 0.1) +
tm_grid(labels.show = TRUE, lines = FALSE)
```

A second approach would be to union the circles and then take the difference:

```{r chunk10}
circle_union_sf <- circles_sf |> st_union(by_feature = FALSE)

pts_outside_circle_union_sf <- pts_sf |> 
  st_difference(circle_union_sf)

tm_shape(circle_union_sf) +
  tm_fill() +
tm_shape(pts_outside_circle_union_sf) +
  tm_dots(col = "orange", size = 0.15) +
tm_grid(labels.show = TRUE, lines = FALSE)
```

\

## CHALLENGE: Plot the points that lie within 0.25 map units of a circle, but are not contained within the circle

One approach we can use to solve this problem is to feed the point layer into `st_is_within_distance()` (which returns a matrix), then feed that matrix into `rowSums()` (which adds up the rows converting TRUE to 1 and FALSE to 0), and then feed the row sums into a condition check (row sum > 0). This will give us a logical vector (TRUE / FALSE) values that indicate whether each point is close to a circle.

Unfortunately `st_is_within_distance()` also returns TRUE when a point is *within* the circle (because distance = 0). Hence we'll need to remove the points within the circles from our set. We can do this by creating a logical vector for intersects, and then write a logical expression that uses the 'and' and 'not' operators to find the points we want.

[Answer](https://bit.ly/3xsXgyR)

```{r chunk11}
library(magrittr)

## Test whether points are near a circle. Note st_is_within_distance() also returns
## TRUE for points that are within the circle (because the distance = 0)
near_or_within_any_circle_yn <- pts_sf |> 
  st_is_within_distance(circles_sf, dist = 0.25, sparse = FALSE) |> 
  rowSums() |> 
  is_greater_than(0)    ## you could also use {. > 0}

## Test the points that are within circles (so we can remove them)
within_any_circle_yn <- pts_sf |> 
  st_intersects(circles_sf, sparse = FALSE) |> 
  rowSums() |> 
  is_greater_than(0)    ## you could also use {. > 0}

## Use the logical 'and' operator (&) and 'not' operator (!) to get those
## rows near but not within circles
near_but_not_within_yn <- near_or_within_any_circle_yn & !within_any_circle_yn

## Count the TRUEs
sum(near_but_not_within_yn)

## Plot to make sure
tm_shape(circles_sf) +
  tm_borders() +
  tm_text("circle_id") +
tm_shape(pts_sf |> filter(near_but_not_within_yn)) +
  tm_dots(col = "brown", size = 0.1) +
tm_grid(labels.show = TRUE, lines = FALSE)
```

Another approach we can take is to create a 'donut ring' around the union  of the threes circles, and then use `st_intersecs()` or `st_intersection()` to plot the points we want. You can make a donut ring with a combination of `st_buffer()` and `st_difference()`.

```{r chunk12}
## Merge the circles into a single multipolygon feature
circles_merged_sf <- circles_sf |> st_union()

## Create the donut ring:
circles_ring_sf <- circles_merged_sf |> 
  st_buffer(dist = 0.25) |> 
  st_difference(circles_merged_sf)
  
## Plot to make sure
tm_shape(circles_ring_sf) +
  tm_polygons()

## Find and plot the points that fall within the donut ring
pts_ring_sf <- pts_sf |> st_intersection(circles_ring_sf)

## Count them
nrow(pts_ring_sf)

## Plot to verify
tm_shape(circles_sf) +
  tm_borders() +
  tm_text("circle_id") +
tm_shape(pts_ring_sf) +
  tm_dots(col = "purple", size = 0.1) +
tm_grid(labels.show = TRUE, lines = FALSE)

```


\

# Plot the Points of Interest that Fall within the Upper Merced Subbasin

Next, we'll apply what we learned to find the Yosemite Points-of-Interest that fall within the Upper Merced HUB-8 Subbasin.

## Import the Watersheds

Start by importing the planning watershed units from [calw221](https://frap.fire.ca.gov/mapping/gis-data/){target="_blank" rel="noopener"}:

```{r chunk13}
## Import the planning watersheds
gpkg_watershd_fn <- "./data/yose_watersheds.gpkg"
yose_watersheds_sf <- st_read(gpkg_watershd_fn, layer="calw221") 

## View attribute table
yose_watersheds_sf |> st_drop_geometry() |> slice(1:6)

## Plot results
tmap_mode("plot")
tm_shape(yose_watersheds_sf) + 
  tm_polygons("MAP_COLORS", palette = "Pastel1") 
```

Note the keyword `MAP_COLORS` tells tmap to select colors at random such that adjacent polygons have different colors.

\

## Lump the Planning Watersheds into HUC-8 Subbasins

Next we'll group the little planning watersheds into bigger "HUC-8" subbasins. This is easy because there is a column for the HUC 8 id number (`HUC_8`) and name (`HUC_8_NAME`).

```{r chunk14}
yose_huc8_sf <- yose_watersheds_sf |> 
  group_by(HUC_8) |> 
  summarise(HUC_8_NAME = first(HUC_8_NAME), num_pws = n())

yose_huc8_sf
```

A convenient feature of `group_by()` is that when applied to a simple feature data frame it will also spatially aggregate (i.e., union) the features based on common values in the grouping column. Plot to verify:

```{r chunk15}
epsg_utm11n_nad83 <- 26911
yose_bnd_utm <- st_read(dsn="./data", layer="yose_boundary") |> 
  st_transform(epsg_utm11n_nad83)

tm_shape(yose_huc8_sf) + 
  tm_polygons("MAP_COLORS", palette = "Pastel1") +
  tm_text("HUC_8_NAME", size = 0.7) +
tm_shape(yose_bnd_utm) +
  tm_borders(col = "red", lwd = 2)
```

\

## Extract Upper Merced HUC-8 Subbasin

Next, pull out just the Upper Merced subbasin and save it as a separate object:

```{r chunk16}
## Filter out just the Upper Merced Subbasin
merced_huc8_sf <- yose_huc8_sf |> 
  filter(HUC_8_NAME == "UPPER_MERCED")
merced_huc8_sf
```

\

## Import the Points-of-Interest

Import the Yosemite POIs:

```{r chunk17}
## Import points of interest
yose_poi_utm <- st_read(dsn="./data", layer="yose_poi") |> 
  select(OBJECTID, POINAME, POILABEL, POITYPE)
```

\

## Identify Intersecting Points-of-Interest

Find out which POIs intersect the Upper Merced subbasin with `st_intersects()`:

```{r chunk18}
try(merced_poi <- yose_poi_utm |> st_intersects(merced_huc8_sf))
```

Oh no - **ERROR message**! Spatial querying requires features to be in the same CRS!

\

To fix this, we can project the Merced HUC-8 layer (which is in CA Albers) to match the POIs (which are UTM):

```{r chunk19}
merced_huc8_utm_sf <- merced_huc8_sf |> 
  st_transform(st_crs(yose_poi_utm))
```

\

Try the intersection test again:

```{r chunk20}
yose_poi_merced_mat <- yose_poi_utm |> st_intersects(merced_huc8_utm_sf, sparse=FALSE)
head(yose_poi_merced_mat)
```

\

## CHALLENGE: How many points-of-interest fall within the Upper Merced subbasin?

*Hint 1*: This is equivalent to asking how many TRUE values there are in the first column of `yose_poi_merced_mat`.

*Hint 2*: To get the first column of a matrix `x`, use `x[ , 1]`.

[Answer](https://bit.ly/3rEOnzn)

```{r chunk21}
## How many points of interest fall in the Upper Merced subbasin?

## Option 1: Sum up the values in the column 1 of the matrix
sum(yose_poi_merced_mat[,1])

## Option 2: Compute a frequency table of the values column 1
table(yose_poi_merced_mat[,1])
```

\

## Subset the POIs that fall within the Upper Merced Subbasin

To extract the POIs in the Upper Merced subbasin, we can feed the first column of `yose_poi_merced_mat` into  `filter()` (which expects TRUE/FALSE values):

```{r chunk22}
## Extract the points that intersect the subbasin to a new sf object
merced_poi_utm <- yose_poi_utm |> 
  filter(yose_poi_merced_mat[,1])
```

\

## Plot the Intersection

Plot to visually verify the results:

```{r chunk23}
## Plot
tm_shape(merced_huc8_utm_sf) +
  tm_polygons(col = "khaki") +
tm_shape(yose_poi_utm) +
  tm_dots(size = 0.1, col = "gray30") +
tm_shape(merced_poi_utm) +
  tm_dots(size = 0.1, col = "dodgerblue")
```

\

## CHALLENGE: How many YNP POIs fall witin the Upper Tuolumne HUC-8 subbasin?

[Answer](https://bit.ly/3AdasK6)

```{r chunk24}
uptuol_huc8_utm_sf <- yose_huc8_sf |> 
  filter(HUC_8_NAME == "UPPER_TUOLUMNE") |> 
  st_transform(st_crs(yose_poi_utm))

## Plot to verify
tm_shape(uptuol_huc8_utm_sf) + 
  tm_polygons() +
tm_shape(yose_poi_utm) +
  tm_dots()

## Count the intersecting points
yose_poi_utm |> 
  st_intersection(uptuol_huc8_utm_sf) |> 
  st_drop_geometry() |> 
  count()

```


## End

Congratulations, you've completed another Notebook! 

To view your Notebook at HTML, save it (again), then click the 'Preview' button in the RStudio toolbar.

