---
title: "Homework 1 - ANSWERS"
author: "Spatial Analysis with R Part 1: The Basics"
output: html_document
---

```{css echo = FALSE}
h2 {
  font-size:18px;
  font-weight:bold;
}
h4.author {
  font-weight:bold;
  color:#666;
  font-style:italic;
}
```

\

## Q1. Create a numeric vector called `x` with 1000 numbers. You can use any function or operator of your choice.

```{r}
x = rnorm(1000)
```

\

## Q2. Use `length()` to verify that `x` has exactly 1000 elements.

```{r}
length(x)
```

\
 
## Q3. Compute the mean, minimum, maximum and range of x.

```{r}
mean(x)
min(x)
max(x)
range(x)
```

\

## Q4. Create a vector called `y` that contains all the odd numbers from 1 to 100.
## HINT: There are several ways to do this. If you get stuck feel free to ask Google for help.

```{r}
y = seq(1, to = 100, by = 2)
```
Or:

```{r}
y = (1:50 * 2) - 1
```

\

## Q5. Write an expression that returns a vector containing the first 10 uppercase letters of the alphabet.

LETTERS[1:10]

\

## Q6. Write an expression that returns a vector containing the first 10 lowercase letters of the alphabet.

letters[1:10]

\

## Q7. If you uncomment the following two lines and run them, you'll get an error. Fix it.

```{r}
# z = 1:5
# z[TRUE,TRUE,FALSE,FALSE,TRUE]
```

\

```{r}
z = 1:5
z[c(TRUE,TRUE,FALSE,FALSE,TRUE)]
```

\

## Q8. Generate 50 random numbers using a function like `runif()`, `rnorm()`, or `sample()`. Save the results to a object called `rand_nums`.

```{r include=FALSE}
set.seed(1234567)
```


```{r}
rand_nums = runif(50)
```

\

## Q9. Find the mean of rand_nums.

```{r}
mean(rand_nums)
```

\

## Q10. Write an expression that returns the elements of rand_nums that are bigger than the average.

```{r}
rand_nums[rand_nums > 0.5690193]
```

\

## Q11. Write an expression that returns the number of elements of rand_nums that are bigger than the average.

```{r}
sum(rand_nums > 0.5690193)
```

