---
title: "Homework 2 - ANSWERS"
author: "Spatial Analysis with R Part 1: The Basics"
output: html_document
---

```{css echo = FALSE}
h2 {
  font-size:18px;
  font-weight:bold;
}
h4.author {
  font-weight:bold;
  color:#666;
  font-style:italic;
}
```

\


## Q1. Load the `maps` package into memory.

```{r}
library(maps)
```

\

## Q2. The maps package comes with a dataset called `world.cities`. What kind of object is 'world.cities'?
  
```{r}
class(world.cities)
```

\

## Q3. How many rows and columns does `world.cities` have?

```{r}
dim(world.cities)
```

Or:

```{r}
nrow(world.cities)
ncol(world.cities)
```

\

## Q4. What is the original data source of world.cities? Hint: datasets have help pages too!

```{r eval=FALSE}
?world.cities
```

*"The data were originally obtained from Stefan Helders' website (http://www.world-gazetteer.com), which now redirects to http://www.populationmondiale.com."*

\

## Q5. What is the average population of the world cities?

```{r}
mean(world.cities$pop)
```

\

## Q6. How many world cities have a population larger than 1,000,000?


```{r}
sum(world.cities$pop >= 1000000)
```

\

## Q7. The maps package also has a dataset called `us.cities`, which includes columns for longitude and latitude. 
Make a scatter plot of the longitude and latitude
## coordinates.

```{r}
plot(us.cities$long, us.cities$lat, pch = 16, asp = 1, cex = 0.5)
```

\

## Q8. Import the following CSV file as a data frame:
## https://raw.githubusercontent.com/ajlyons/rspatial_bgs23/main/notebooks/data/sf_libraries.csv
## How many libraries are there in San Francisco?

```{r}
libraries <- read.csv("https://raw.githubusercontent.com/ajlyons/rspatial_bgs23/main/notebooks/data/sf_libraries.csv")

nrow(libraries)
```

\

## Q09. Use `runif()` to generate 20 random numbers uniformly distributed between 100 and 200.

```{r}
runif(20, min = 100, max = 200)
```

\

## Q10. Rewrite the following expression with pipes. You can use either the traditional pipe (`%>%`) from the magrittr package, or the more recent native pipe (`|>`).

<p style="font-family:monospace; font-size:120%;">sum(round(rnorm(n = 200, mean = 50, sd = 5)))</p>

\

```{r}
rnorm(n = 200, mean = 50, sd = 5) |> round() |> sum()
```

or

```{r}
library(magrittr)
rnorm(n = 200, mean = 50, sd = 5) %>% round() %>% sum()
```

\

