---
title: "Homework 3 - ANSWERS"
author: "Spatial Analysis with R Part 1: The Basics"
output: html_document
---

```{css echo = FALSE}
h2 {
  font-size:18px;
  font-weight:bold;
}
h4.author {
  font-weight:bold;
  color:#666;
  font-style:italic;
}
```

## Setup

First we load some data from the San Francisco Open Data Portal. Specifically, we'll import a selection of 311 calls from San Francisco ([more info](https://data.sfgov.org/resource/vw6y-z8j6)).

```{r}
## Define a URL to a CSV
url1 <- "https://data.sfgov.org/resource/vw6y-z8j6.csv?$select=service_request_id,requested_datetime,agency_responsible,service_name,service_subtype,address,neighborhoods_sffind_boundaries,source&$limit=100&$where=requested_datetime between '2023-02-21T00:00:00' and '2023-02-21T23:59:59'"

## Next, read in the data. (URLencode() removes spaces in the URL)
df <- read.csv(URLencode(url1))

## Finally, change the name of one of the columns
library(dplyr)
df <- df |> rename(neighborhood = neighborhoods_sffind_boundaries)
```

\

If you'd like, uncomment the following line to view the data frame in a View pane:

```{r}
# View(df)
```

\

## Q1. How many rows and how many columns does this data frame have?

```{r}
dim(df)
```

\

## Q02. Write an expression that returns the column names of this data frame.

*Hint*: Look for a function that returns the names of a data frame

```{r}
names(df)
```

\

## Q03. Write an expression that returns the data frame sorted by the responsible agency.

```{r}
df |> arrange(agency_responsible) |> head()
```

\

## Q04. Write an expression that just returns three of columns (of your choice).

```{r}
df |> select(service_name, service_subtype, source) |> head()
```

\

## Q05. Write an expression that just returns 311 calls that were submitted by phone.
 
```{r}
df |> filter(source == 'Phone') |> head()
```
 
\

## Q06. How many 311 calls were submitted by phone?

```{r}
df |> filter(source == 'Phone') |> nrow()
```

\

## Q07. Write an expression that returns the number of calls per neighborhood, sorted *alphabetically by neighborhood*.

```{r}
df |> group_by(neighborhood) |> summarise(num_calls = n()) |> arrange(neighborhood)
```

\

## Q08. Write an expression that returns the number of calls per neighborhood, sorted by the *number of calls*.

```{r}
df |> group_by(neighborhood) |> summarise(num_calls = n()) |> arrange(desc(num_calls)) 
```

\

## Q09. What is the most common service requested?

```{r}
df |> group_by(service_name) |> summarize(num_recs = n()) |> arrange(desc(num_recs))
```

Or:

```{r}
df |> group_by(service_name) |> summarize(num_recs = n()) |> slice_max(num_recs, n = 1)
```

\

## Q10. What is the least common service requested?

```{r}
df |> group_by(service_name) |> summarize(num_recs = n()) |> slice_min(num_recs, n = 1)
```





