---
title: "Homework 5 - ANSWERS"
author: "Spatial Analysis with R Part 2"
output: html_document
---

```{css echo = FALSE}
h2 {
  font-size:18px;
  font-weight:bold;
}
h4.author {
  font-weight:bold;
  color:#666;
  font-style:italic;
}
```

## Q01. Import the Yosemite Watersheds

Looking at the datasets page on the course website, we see that the Yosemite watersheds are stored in a GeoPackage. 

Step 1 is to construct the path to the geopackage file. Here we'll use the here() function that returns the path to the RStudio project folder. 

```{r}
yose_water_fn <- here::here("./notebooks/data/yose_watersheds.gpkg")
file.exists(yose_water_fn)
```


\

Note: If you didn't open the RStudio project, `here()` won't work. In that case, you can enter your own path. For example:

`yose_water_fn <- "c:/path/to/data/yose_watersheds.gpkg"`

\

Next, let's view the layers in this GeoPackage:

```{r}
library(sf)
st_layers(yose_water_fn)
```

\

Import `calw221`:

```{r}
yose_water_sf <- st_read(yose_water_fn, layer="calw221")
```

\

## Q02. What coordinate reference system are the watersheds in?

```{r}
st_crs(yose_water_sf)
```

Answer: California Albers / NAD83

\

## Q03. Plot the watersheds

```{r}
plot(yose_water_sf$geom, axes = TRUE)
```


